A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-23 latex/main.tex
Mutation spectra

\subsubsection*{Comparison of mutational spectra}

To validate the mutational information contained in the sc2ts ARG,
we compared the all-site mutational spectra of major VOCs in the ARG,
a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024},
and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
Mutational spectra are defined as the relative proportions of
each possible type of nucleotide change.

First, we calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
in the sc2ts ARG, considering mutations (ignoring indels) above the sample nodes and internal nodes
which have Scorpio labels associated with the VOCs
(e.g., for Alpha, B.1.1.7-like and B.1.1.7-like+E484K).
These Scorpio labels were assigned to the aligned sequences of the nodes in the ARG using pangolin.
Note that $\sim$98\% of all the mutations in the ARG occur
above the nodes that have a Scorpio label associated with these VOCs
(excluding the nodes with unassigned Scorpio labels).

Next, we calculated the mutational spectra of the same VOCs
using the mutational counts from an UShER phylogeny analyzed in Bloom et al.\cite{Bloom2023},
which was built using a Nextclade pipeline and
has samples collected worldwide up to November, 2022.
To obtain the WHO label corresponding to a Nextclade-defined clade, we used the following mapping:
20I, Alpha; 21I and 21J, Delta; 21K, BA.1; 21L, BA.2; 22A, BA.4; and 22B, BA.5.
The mutation counts were downloaded from https://github.com/jbloomlab/SARS2-mut-spectrum/.

Lastly, we calculated the mutational spectra of the same VOCs
in the Viridian UShER phylogeny\cite{Hunt2024},
which was pruned down to the samples present in the sc2ts ARG.
We used mutations above the sample nodes,
for which Scorpio labels are available
in the Viridian metadata provided by Hunt et al.\cite{Hunt2024}.

>>>>

\subsubsection*{Comparison of mutational spectra}
% To validate the mutational information contained in the sc2ts ARG,
% we compared the all-site mutational spectra of major VOCs in the ARG,
% a state-of-the-art UShER phylogeny built using the Viridian dataset\cite{Hunt2024},
% and another UShER phylogeny analyzed in a previous study\cite{Bloom2023}.
% Mutational spectra are defined as the relative proportions of
% each possible type of nucleotide change.

We calculated the mutational spectra of Alpha, Delta, Omicron BA.1, BA.2, BA.4, and BA.5
in the sc2ts ARG, considering mutations (ignoring indels) above the sample 
nodes and internal nodes
which have Scorpio labels associated with the VOCs
(e.g., for Alpha, B.1.1.7-like and B.1.1.7-like+E484K).
Note that $\sim$98\% of all the mutations in the ARG occur
above the nodes that have a Scorpio label associated with these VOCs
(excluding the nodes with unassigned Scorpio labels).

We calculated the mutational spectra of the same VOCs
in the intersection UShER tree in tskit format using a similar approach.
Here, we used the Scorpio sample assignments provided in the Viridian metadata,
and therefore restricted to mutations above sample nodes only.

Next, we calculated the mutational spectra of the same VOCs
using the mutational counts from an UShER phylogeny 
analyzed by Bloom et al.\cite{Bloom2023},
which was built using a Nextclade pipeline and
has samples collected worldwide up to November, 2022.
To obtain the WHO label corresponding to a Nextclade-defined clade, we used the following mapping:
20I, Alpha; 21I and 21J, Delta; 21K, BA.1; 21L, BA.2; 22A, BA.4; and 22B, BA.5.
The mutation counts were downloaded from
\url{https://github.com/jbloomlab/SARS2-mut-spectrum/}.

See
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/fig\_mutational\_spectra.ipynb}{fig\_mutational\_spectra.ipynb}
notebook for details.
